---
title: "Activity Monitoring Data"
keep_md: yes
date: "21 March 2018"
output:
  md_document: default
  html_document: default
---

## Loading and preprocessing the data

```{r, echo = TRUE}
# Load the data assuming it is saved under its original name in the wording directory
setwd("\\\\OAKSBS11/RedirectedFolders/guillaumebonnissent/My Documents")
data <- read.csv("activity.csv", sep = ",", header = TRUE)

# Convert date variable to date format
data$date <- as.Date(data$date, "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

```{r}
# Create new dataset with no missing values (NB: missing values are for all 5-minute interval of one day)
dateswithnodata <- unique(data[is.na(data$steps), ]$date)
data2 <- data[!(data$date %in% dateswithnodata), ]
# Calculate the total number of steps taken per day
stepsperday <- tapply(data2$steps, data2$date, sum)
# Make a histogram of the total number of steps taken each day using ggplot2 package with a class interval of 2,500
library(ggplot2)
qplot(stepsperday, binwidth = 2500, main = "Distribution of Total Number of Steps Taken per Day",
      xlab = "Total Number of Steps Taken per Day", ylab = "Frequency") + theme(plot.title = element_text(hjust = 0.5))
```

```{r}
# Calculate the mean of the total number of steps taken per day
stepsmean <- mean(stepsperday)
# Calculate the median of the total number of steps taken per day
stepsmedian <- median(stepsperday)
```

The mean of the total number of steps taken per day is `r stepsmean`, and its median is `r stepsmedian`.

## What is the average daily activity pattern?

```{r}
# Calculate the average number of steps taken per 5-minute interval averaged across all days
stepsmeanper5min <- tapply(data$steps, data$interval, mean, na.rm = TRUE)
# Make a time series plot of the 5-minute interval (x-axis) by the average number of steps averaged across all days (y-axis)
plot(x = names(stepsmeanper5min), y = stepsmeanper5min, type = "l",
     main = "Average Number of Steps Taken per 5-minute Interval",
     xlab = "5-minute Interval", ylab = "Average Number of Steps Taken")
```

```{r}
# Calculate the 5-minute interval with the maximum average number of steps taken
maxsteps5min <- which.max(stepsmeanper5min)
max5min <- names(maxsteps5min)
```

The 5-minute interval with the maximum average number of steps taken is interval `r max5min`, during which an average of `r maxsteps5min` steps were taken.

## Imputing missing values

```{r}
# Calculate the total number of missing values in the dataset
missingsteps <- sum(is.na(data$steps))
```

The total number of missing values in the dataset is `r missingsteps`. A strategy to replace those missing values is to replace them with the total average number of steps taken per day but apportioned by 5-minute interval according to the average number of steps taken by 5-minute intervals.

```{r}
# Calculate the average number of steps taken by 5-minute intervals for one day which would average the total average number of steps taken per day
stepsmeanper5min2 <- stepsmeanper5min / sum(stepsmeanper5min) * stepsmean
# Create a new dataset equal to the original dataset but with the missing values filled in
data3 <- data
for(i in 1:nrow(data3)) {
          if(is.na(data3$steps[i])) {
                    data3$steps[i] <- stepsmeanper5min2[names(stepsmeanper5min2) == data3$interval[i]]
          }
}
if(any(is.na(data3$steps))) {
          paste("There are still", sum(is.na(data2$steps)), "missing values")
}
```

```{r}
# Calculate the total number of steps taken per day on the new dataset
stepsperday2 <- tapply(data3$steps, data3$date, sum)
# Make a histogram of the total number of steps taken each day using ggplot2 package with a class interval of 2,500 from the new dataset
library(ggplot2)
qplot(stepsperday2, binwidth = 2500, main = "Distribution of Total Number of Steps Taken per Day (missing values replaced)",
      xlab = "Total Number of Steps Taken per Day", ylab = "Frequency") + theme(plot.title = element_text(hjust = 0.5))
```

```{r}
# Calculate the mean of the total number of steps taken per day on the new dataset
stepsmean2 <- mean(stepsperday2)
# Calculate the median of the total number of steps taken per day on the new dataset
stepsmedian2 <- median(stepsperday2, na.rm = TRUE)
```

As expected considering our strategy to replace missing values, the mean of the total number of steps taken per day on the dataset with the missing values replaced is the same (`r stepsmean2` against `r stepsmean` on the original dataset), and its median hasn't materially changed (`r stepsmedian2` against `r stepsmedian`).

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# Create a new factor variable in the dataset with two levels indicating whether a given date is a week day or weekend day in the dataset
# with missing values filled in
data3$day <- ifelse(weekdays(data3$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
# Calculate the average number of steps taken per 5-minute interval and type of day, averaged across all days 
databyweekday <- with(data3, aggregate(steps, by = list(interval, day), FUN = mean))
names(databyweekday) <- c("interval", "day", "steps")
# Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
ggplot(databyweekday, aes(interval, steps)) + geom_line() + facet_wrap(~ day, nrow = 2) + xlab("Interval") + ylab("Number of steps")
```

